---
title: "cfbfastR NCAA Football Coverage Audit: DII and DIII"
subtitle: "Tartan D3 Football Scraper -- Data Availability Checks"
author: "Sam"
date: "September 7, 2026"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Purpose

Before building a scraper, this document establishes whether the existing
`cfbfastR` ecosystem already provides NCAA Division II or Division III football
data. `cfbfastR` version 3.x added a `load_ncaa_mfb_*()` loader family sourced
from stats.ncaa.org, documented as covering "FCS and lower divisions." These
checks test that claim directly against the released data.

The verdicts below are computed from the data, not asserted, so this audit
stays valid if coverage changes in a future release. Re-run to re-check.

NCAA football division codes on stats.ncaa.org:

| Code | Division            |
|-----:|:--------------------|
| 11   | FBS (Division I-A)  |
| 12   | FCS (Division I-AA) |
| 2    | Division II         |
| 3    | Division III        |

## Setup

```{r libraries}
# install.packages("cfbfastR")
# or: remotes::install_github("sportsdataverse/cfbfastR")
stopifnot(requireNamespace("cfbfastR", quietly = TRUE))

library(cfbfastR)
library(dplyr)
library(tidyr)

# Normalize the division field to a clean integer-as-character key,
# robust to whether the loader returns it as int, double, or string.
norm_div <- function(x) as.character(as.integer(round(as.numeric(x))))
div_levels <- c("11", "12", "2", "3")
div_labels <- c("11" = "FBS", "12" = "FCS", "2" = "DII", "3" = "DIII")
```

## 1. Available NCAA football loaders

Confirm which loaders the installed version exposes before calling them. If a
function name below differs from what later chunks use, the package API has
shifted and the calls need updating.

```{r loaders}
ls("package:cfbfastR", pattern = "ncaa_mfb")
```

## 2. Division codes present in the team universe

Load the team universe once (every published season) and reuse it throughout.
The teams dataset is the ceiling on coverage: no game, drive, or play can exist
for a team absent here. First, the raw division codes actually present:

```{r teams-load}
teams <- load_ncaa_mfb_teams(seasons = TRUE)

sort(unique(teams$division))
```

The codes are not self-documenting, so anchor their meaning to teams whose
division is already known. Sampling teams under each code from the most recent
season makes the mapping obvious (recognizable FBS vs FCS programs).

```{r division-sample}
teams |>
  filter(season == max(season)) |>
  group_by(division) |>
  slice_head(n = 8) |>
  select(division, team_name) |>
  print(n = Inf)
```

## 3. Team universe by division, every published season

Counting by division across all seasons gives the whole-picture view. Columns
are forced to include all four divisions so any zero is visible rather than
silently dropped.

```{r teams-coverage}
coverage <- teams |>
  mutate(division = norm_div(division)) |>
  filter(division %in% div_levels) |>
  count(season, division) |>
  complete(season, division = div_levels, fill = list(n = 0)) |>
  mutate(division = factor(div_labels[division], levels = unname(div_labels))) |>
  pivot_wider(names_from = division, values_from = n, values_fill = 0) |>
  arrange(season)

# Team count by division and season
print(coverage, n = Inf)
```

## 4. Verdict: does DII or DIII exist at all?

```{r verdict}
present_codes <- teams |> distinct(division) |> pull(division) |> norm_div()

verdict <- tibble(
  division      = c("DII", "DIII"),
  code          = c("2", "3"),
  team_seasons  = c(
    sum(norm_div(teams$division) == "2"),
    sum(norm_div(teams$division) == "3")
  ),
  present       = c("2", "3") %in% present_codes
)

# DII / DIII presence in the cfbfastR team universe
print(verdict)

has_d2 <- "2" %in% present_codes
has_d3 <- "3" %in% present_codes
```

**DII present in cfbfastR:** `r has_d2`  
**DIII present in cfbfastR:** `r has_d3`

```{r verdict-prose, echo=FALSE, results='asis'}
if (!has_d2 && !has_d3) {
  cat("Neither Division II nor Division III appears in any published season. ",
      "cfbfastR covers Division I only (FBS and FCS). Both lower divisions ",
      "are an open gap.", sep = "")
} else {
  cat("At least one lower division is present. Review the coverage table ",
      "above for completeness before assuming the gap is closed.", sep = "")
}
```

## 5. Unexpected division codes

Guard against the four-code assumption. Any code outside {11, 12, 2, 3} would
surface here and warrant investigation.

```{r unexpected-codes}
unexpected <- teams |>
  mutate(division = norm_div(division)) |>
  filter(!division %in% div_levels) |>
  count(division, name = "team_seasons")

print(unexpected)
```

## 6. Game-level confirmation

The team-level result should hold at the game level, since the schedule dataset
is keyed on the same team universe. Named DII and DIII programs should return
zero rows in the most recent season's schedule.

```{r schedule-check}
audit_season <- max(teams$season)

sched <- load_ncaa_mfb_schedule(seasons = audit_season)

d3_programs <- c("Carnegie Mellon", "Johns Hopkins", "Mount Union",
                 "North Central (IL)")
d2_programs <- c("Ferris State", "Grand Valley St.", "Colorado Mines")

game_check <- tibble(
  division = c("DIII", "DII"),
  programs_tested = c(length(d3_programs), length(d2_programs)),
  games_found = c(
    sched |> filter(team_name %in% d3_programs |
                      opponent %in% d3_programs) |> nrow(),
    sched |> filter(team_name %in% d2_programs |
                      opponent %in% d2_programs) |> nrow()
  )
)

# Named lower-division programs found in the most recent schedule
cat("Schedule season audited:", audit_season, "\n")
print(game_check)
```

Note: a zero here is the expected result given a Division I-only universe, not
a name-matching failure. Program name strings on stats.ncaa.org may differ from
those above, but any nonzero count would still be meaningful.

## 7. Optional: play-by-play level

Not run by default, since the play-by-play dataset is large and the team
universe already bounds coverage. Enable only to confirm directly.

```{r pbp-check, eval=FALSE}
pbp <- load_ncaa_mfb_pbp(seasons = max(teams$season))
# Inspect whatever division/team identifiers the pbp schema carries and
# confirm no DII/DIII teams appear.
glimpse(pbp)
```

## 8. Summary

```{r summary, echo=FALSE, results='asis'}
seasons_span <- paste0(min(teams$season), " to ", max(teams$season))
cat(sprintf(
  paste0("- Seasons audited: %s\n",
         "- Divisions present: %s\n",
         "- DII available in cfbfastR: **%s**\n",
         "- DIII available in cfbfastR: **%s**\n"),
  seasons_span,
  paste(sort(unique(norm_div(teams$division))), collapse = ", "),
  has_d2, has_d3
))
```

If DII and DIII both read `FALSE`, the scraper project targets a genuine,
complete gap in the SportsDataverse ecosystem, and stats.ncaa.org (which does
carry both divisions under codes 2 and 3) is the source to build against.

## Reproducibility

```{r session}
sessionInfo()
```
